Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-01-07 demystify.tex
updates

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Common concerns include (i) the \textbf{difficulty of getting admitted}, (ii) the \textbf{cost of graduate study} in the US, and for some, the fear of \textbf{discrimination} e.g., based on gender identity or cultural background.
Much of these uncertainties stem from the lack of information and guidance on the admission process in the US (social places like Facebook and Reddit are full of confusion and contradictory information).  This book aims to address these concerns.

@@

This book also aim to explain the \emph{``why''} behind the admission process, e.g., why LoRs are important, why you should not draft your own LoR, why you should contact professors, etc. There are numerous \href{https://github.com/nguyenthanhvuh/phd-cs-us}{resources online} from students  that tell you \emph{``what''} to do, but few explain the \emph{``why''}. Understanding the reason and mindset of the admission committee and profs. can help you prepare better.

@@

Some universities do waive, e.g., Rice does for all PhD applicants and many universities for domestic students (\autoref{sec:domestic-students}), however most do not for international students.
Some require proof of financial difficulties, e.g., a statement from a financial adviser or a bank statement.

@@

As mentioned in~\autoref{sec:lor-writers}, LoR writers should be someone who knows you well and has the credibility to evaluate your research ability.  In the US, students usually explicitly ask if the writer would be willing to provide a strong letter, and the writer are also very direct in their response.  If they are not willing (you should be thankful that they are honest with you), then you should ask someone else.

Below are some tips to approach your LoR writers:
\begin{itemize}
\item \textbf{Ask in advance} You should ask for LoRs \emph{at least a month} before the deadline. People are busy and writing a strong LoR takes time (\autoref{sec:lor-writing}), so give them enough time to write a strong letter for you.

@@

\item \textbf{Help your writers (\autoref{sec:help-your-LOR-writers})}: You should tell your writers the programs you are applying to, their deadlines, etc.  You can also share your SOP with them and other details about your research experience and potential.

\item \textbf{Ask for feedback} If the writer is very close to you, you can ask them for feedback on your SOP (\autoref{chap:sop}) and other application materials. If the writer is a professor, they might have served in adcom committees, seen many SOPs, and can provide valuable feedback.

@@

After your writers have submitted all of their letters, \emph{thank them} (\autoref{sec:thank-writers}).  Finally, let them know the outcome of your applications and stay in touch with them.  This will help you build a relationship with them and you might need them to write for you again in the future.
\end{itemize}

@@

Here we look at publications and other research experiences that can strengthen your application.
\autoref{chap:research-opportunities} provides more information on how to find research opportunities (e.g., during your undergrad study).

@@

Publications are never required for application, however given the competitiveness of CS admission, they can significantly strengthen your application and are becoming the norm. Applicants admitted to top schools, especially in popular fields such as ML and NLP, often have multiple first-authored papers at top places.

\paragraph{What if I don't have any pubs} It is understandable that many students do not have the opportunity to publish papers. Thus, other writings, even those under submissions or rejected, would still be good and better than nothing.  Be sure to upload your these with your application (\autoref{sec:writing-sample}) and talk about them in your SOP (\autoref{chap:sop}).

@@

\section{Work Experience} Work experience at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Note that the emphasis here is \emph{research} places, not software development or non-research work. For example, working at a FAANG (Facebook, Amazon, Apple, Netflix, Google) company as a software engineer is not research experience, though it can be helpful for MS applications discussed in \autoref{chap:ms}. Similarly, a LoR from your supervisor for non-research experience might not count much. So do not spend much time talking about non-research job in your SOP.

Note that adcom reviewers might not be familiar with all research labs, especially those outside the US. For example, while VinAI is  well-known in Vietnam and potentially in Asia, it might not be very well-known in the US. So you or your LoR writers should explicitly say something about them in your statement.  In general, if you did some good research work, then you should mention that in your SOP and ask your supervisor to write about it in their LOR (\autoref{chap:lor}).

% \begin{commentbox}[Hung:]
%   The reputation of VinAI has been increasing steadily over the past few years; many of my colleagues heard about VinAI.
% \end{commentbox}

\section{Competitions} Winning \emph{internationally recognized competitions} can demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.  
So do talk about them in your SOP (\autoref{chap:sop}) and have your LoR writers mention them in their letters (\autoref{sec:help-your-LOR-writers}).

@@

\paragraph{Not a Good fit and Bad luck}  You could have an excellent profile (e.g., strong research and LoRs), but if you are interested in a research area that the program does not have, you will not be admitted.
Similarly, if no faculty is willing to advise you (e.g., they are on sabbatical or personal leave, do not have sufficient funding, or already have too many students), you will not be admitted.  This is actually good for you, as you don't want to be in a program where no one can advise you.

@@

\begin{itemize}
    \item \textbf{Send emails to the wrong prof.}. Most students often do not pay attention or know about this, but a very common reason why you don't get a reply is that you send email to profs. who do not / cannot advise CS PhD students (e.g., adjunct, emeritus, teaching, non-CS, etc.). See \autoref{sec:faculty-types} for details. While these profs. might be able to co-advise, they typically do not have the bandwidth, funding, and the desire to mentor students for research (they are already overloaded with teaching and other service duties). So do some homework before sending emails, e.g., most CS profs. who are active in research will have a website with their work (publications) and students; some also have lab website dedicated to their group. Most teaching and adjunct profs. do not have such a research page.

    \item \textbf{Generic}. You should already know this! A copy and paste kind of email or those that can be sent to multiple ppl with very little modifications show the lack of interest and will be treated as spam. Most likely we will not reply to these emails.

@@

    \item \textbf{Flowery greetings and language}. Don't use ``Dear esteemed professor'', don't ``hey there''.  Do not call the prof. by their first name in the first email (some don't care but you don't want to take the risk -- you don't know them that well yet).  Do not use Mr., Mrs., etc. To be safe, use Prof. Lastname or Dr. Lastname (\autoref{sec:address}).

@@

    \item \textbf{Fancy format}.  Do not use colors, fancy fonts or formats. While not really a kiss of death, it is very annoying, especially for ppl in CS (and probably many other fields) who often prefer plain messages.

@@

%\section{"Chance me" questions}\label{sec:chance-me}

%Reddit and other grad admission forums are full of ``Chance me'' and ``roast my profile'' questions. These are questions where students post their profiles and ask others to evaluate their chances of getting admitted to their choice of schools.

%These questions are not useful and should be avoided. First, no one knows! Even adcom members can't tell if you would get admitted to their programs or not. As mentioned throughout this guide, the admission process is complex and involves many factors that are not part of your profile, e.g., whether the faculty have funding or are interested in your background. Second, answers to this questions are often not helpful to you.  Faculty and people knowledgeable about the process will not answer these questions, and those who do are often not knowledgeable and might give you false hope or despair. OK, someone say your GRE is low (which does not even matter for most schools (\autoref{chap:standard-tests}) , are you going to retake it?  Or someone says your GPA is low, are you going to apply to a different school?  No, you should apply to the schools you want to go to and see what happens.  If you don't get in, then you can ask for feedback and improve your profile for the next round.

\section{``Chance me'' on getting admitted to Top 10}\label{sec:chance-me}

Many students, especially international students, aim for very top CS PhD programs such as Stanford and MIT, and Ivy Leagues. Every year Reddit and other forums have numerous students asking for evaluation of their chances of getting into these schools (the so called \emph{``chance me''} or \emph{``roast my CV/profile''} posts) and then later posts on being \emph{``rejected everywhere''}.

\paragraph{You are unlikely to get in these schools}  While being ambitious is always good, it is important to be realistic, and the harsh reality is that it is very unlikely that you will get in MIT, or Harvard. Why? people who do get in these schools often were explicitly encouraged by their mentors and LoR writers, who themselves might be alumni of these schools or are academic celebrities who've sent many of their students there. If you are not in this category (doubt you are, because you're asking random people on the internet instead of your trusted LoR writers), then you're likely not going to get in, and MIT and Stanford would glady take your application money.

@@

@@

\section{How Applications are Evaluated}

\begin{itemize}
\item Each applicant is assigned to ~3 reviewers who have expertise in the applicant's research topics (so only your field but also your specific area).  This means reviewers can be quite knowledgeable in your field and your work, your LoR writers and their work, and thus can evaluate your application quite well.

@@

\section{Intellectual Merit and Broader Impacts}
Your will be evaluated based on two criteria: Intellectual Merit (IM) and Broader Impacts (BI).

@@

\section{Common Pitfalls}

>>>>

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Common concerns include (i) the \textbf{difficulty of getting admitted}, (ii) the \textbf{cost of graduate study} in the US, and for some, the fear of \textbf{discrimination} e.g., based on gender identity or cultural background.

Much of these uncertainties stem from the lack of information and guidance on the admission process in the US. Social platforms like Facebook and Reddit are full of confusion and contradictory information.
This book aims to address these concerns.

@@

This book also aims to explain the reasons behind the admission process. There are numerous \href{https://github.com/nguyenthanhvuh/phd-cs-us}{resources online} from students  that tell you \emph{``what''} to do, but few explain the \emph{``why''}, e.g., why LoRs matter so much, why you should not draft your own LoR, why you should contact professors, etc. Understanding the reason and mindset of the admission committee and profs. can help you prepare better.

@@

Some universities do waive, e.g., Rice, TTIC do not have fee for PhD applications and many universities for domestic students (\autoref{sec:domestic-students}), however most do not for international students.
Some require proof of financial difficulties, e.g., a statement from a financial adviser or a bank statement.

@@

As mentioned in~\autoref{sec:lor-writers}, LoR writers should be someone who knows you well and has the credibility to evaluate your research ability.  In the US, it's common for students to explicitly ask if the writer would be willing to provide a strong letter, and the writer are also very direct in their response.  If they are not willing (you should be thankful that they are honest with you), then you should ask someone else.

Below are some tips to approach your LoR writers:
\begin{itemize}
\item \textbf{Ask in advance} You should ask for LoRs \emph{at least a month} before the deadline. People have commitments and writing a strong LoR takes time (\autoref{sec:lor-writing}), so give them enough time to write a strong letter for you.

@@

\item \textbf{Help your writers (\autoref{sec:help-your-LOR-writers})}You should tell your writers the programs you are applying to, their deadlines, etc.  You can also share your SOP with them and other details about your research experience and potential.

\item \textbf{Ask for feedback} If the writer is very close to you and willing to, you can ask them for feedback on your SOP (\autoref{chap:sop}) and other application materials. If the writer is a professor, they might have served in adcom committees, seen many SOPs, and can provide valuable feedback.

@@

After your writers have submitted all of their letters, \emph{thank them} (\autoref{sec:thank-writers}).  Let them know the outcome of your applications and stay in touch with them.  This will help you build a relationship with them and you might need them to write for you again in the future.
\end{itemize}

@@

Here we look at publications and other research experiences that can strengthen your application.
\autoref{chap:research-opportunities} provides more information on how to find research opportunities, e.g., during your undergrad study.

@@

Publications are \emph{never required} for application, however given the competitiveness of CS admission, they can significantly strengthen your application and \emph{are becoming the norm}. Applicants admitted to top schools, especially in popular fields such as ML and NLP, often have multiple first-authored papers at top places. Fig.~\ref{fig:wu} shows examples of applicants to Stanford CS PhD.

\begin{figure}
\begin{center}
    \includegraphics[scale=0.3]{files/wu.jpeg}
\end{center}
\caption{Applicants to top CS programs have multiple first-authored papers at top places.}\label{fig:wu}
\end{figure}

\paragraph{What if I don't have any pubs} It is understandable that many students do not have the opportunity to publish papers. Thus, other writings, even those under submissions or even rejected, would still be good and better than nothing. 
Be sure to upload your these with your application (\autoref{sec:writing-sample}) and talk about them in your SOP (\autoref{chap:sop}). Adcom members can quickly skim over the paper to determine its quality (\autoref{sec:ievaluate}).

@@

\section{Work Experience} Work experience at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  The emphasis here is \emph{research} places, not software development or non-research work. For example, working at a FAANG (Facebook, Amazon, Apple, Netflix, Google) company as a software engineer does not count as research experience, although it can be helpful for MS applications discussed in \autoref{chap:ms}. Similarly, a LoR from your supervisor for non-research experience might not count much. So do not spend much time talking about development job in your SOP.

Note that adcom reviewers might not be familiar with all research labs, especially those outside the US. For example, while VinAI is well-known in Vietnam and potentially in Asia, it might not be very well-known in the US. So you or your LoR writers should explicitly say something about them in your statement or letter.  In general, if you did some good research work, then you should mention that in your SOP and ask your supervisor to write about it in their LOR (\autoref{chap:lor}).

\section{Competitions}\label{sec:competitions}

Winning \emph{internationally recognized competitions} can demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.  
So do talk about them in your SOP (\autoref{chap:sop}) and have your LoR writers mention them in their letters (\autoref{sec:help-your-LOR-writers}).

@@

\paragraph{Not a Good Fit and Bad luck}  You could have an excellent profile (e.g., strong research and LoRs), but if you are interested in a research area that the program does not have, you will not be admitted.
Similarly, if no faculty is willing to advise you (e.g., they are on sabbatical or personal leave, do not have sufficient funding, or already have too many students), you will not be admitted.  This is actually good for you, as you don't want to be in a program where no one can advise you.

@@

\begin{itemize}
    \item \textbf{Send emails to the wrong prof.}. Many students do not pay attention or know about this, but a very common reason why you don't get a reply is that you send email to profs. who do not / cannot advise CS PhD students (e.g., adjunct, emeritus, teaching, non-CS). See \autoref{sec:faculty-types} for details. While these profs. might be able to co-advise, they typically do not have the bandwidth, funding, and the desire to mentor students for research (they are already overloaded with teaching and other service duties).

    So do some homework before sending emails, e.g., most CS profs. who are active in research will have a website with their work (publications) and students; some also have lab website dedicated to their group. Most teaching and adjunct profs. do not have a research group or students listed on their website.

    \item \textbf{Generic}. You should already know this! A copy and paste kind of email or those that can be sent to multiple people with very little modifications show the lack of interest and will be treated as spam. Most likely we will not reply to these emails.

@@

    \item \textbf{Flowery greetings and language}. Don't use ``Dear esteemed professor''.  Do not call the prof. by their first name in the first email (some don't care but you don't want to take the risk -- you don't know them that well yet).  Do not use Mr., Mrs., etc. To be safe, use Prof. Lastname or Dr. Lastname (\autoref{sec:address}).

@@

    \item \textbf{Fancy format}.  Do not use colors, fancy fonts or formats. While not really a kiss of death, it is very annoying, especially for people in CS (and probably many other fields) who often prefer plain messages.

@@

\section{``Chance me'' on getting admitted to T10 programs}\label{sec:chance-me}

Many students, especially international students, aim for very top schools such as Stanford and MIT, and Ivy Leagues. Every year Reddit and other forums have numerous students asking for evaluation of their chances of getting into these schools (the so called \emph{``chance me''} or \emph{``roast my CV/profile''} posts) and then later posts on being \emph{``rejected everywhere''}.

\paragraph{You are unlikely to get in these schools}  While being ambitious is always good, it is important to be realistic, and the harsh reality is that it is very unlikely that you will get in MIT or Harvard unless you are very exceptional (in which case you would not be asking for chances on Reddit).

People who do get in these schools often were explicitly encouraged by their mentors and LoR writers, who themselves might be alumni of these schools or are academic celebrities who've sent many of their students there. If you are not in this category (doubt you are, because you're asking random people on the internet instead of your trusted LoR writers), then you're likely not going to get in, and MIT and Stanford would glady take your application money.

@@

\section{Checking for AI in SOPs}\label{sec:ai-sop}

As AI/LLMs become more popular, many students wonder if they could use it to help with their statements and university/adcom reviewers would check and penalize AI-assisted writing.

While I cannot speak for others, I do not check your statements for AI contents. First, I do not have the time to do that.  It is much easier for me to just read the statement and see if it makes sense and stands out (\autoref{sec:ievaluate}). Hint: AI-generated content reads very strangely and faculty is just too experienced in reading essays and SOPs from students
to not notice it.
Second, AI-checking technology is very unreliable and inconsistent (e.g., a checker might claim that 80\% of an essay is AI-generated while another says it is 0\%).  Third, while it is not OK to use AI to write your statement (as it is both easy to detect and just feels wrong), I think it is fine to use it to help you polish your writing, e.g., I find the ``proofread'' feature in Apple's \texttt{Writing Tools} useful to fix writing mistakes.

@@

\section{NSF GRFP}\label{sec:nsf-grfp}

\subsection{How Applications are Evaluated}

\begin{itemize}
\item Each applicant is assigned to ~3 reviewers who have expertise in the applicant's research topics (so not only your field but also your specific area).  This means reviewers can be quite knowledgeable in your field and your work, your LoR writers and their work, and thus can evaluate your application quite well.

@@

\subsection{Intellectual Merit and Broader Impacts}
Your will be evaluated based on two criteria: Intellectual Merit (IM) and Broader Impacts (BI).

@@

\subsection{Common Pitfalls}
